---
title: "Lab: Creating Good Graphics"
author: "Grant Bauman"
format: html
number-sections: true
number-depth: 2
messages: false
errors: false
warnings: false
---

::: callout
You can see the purpose of this assignment as well as the skills and knowledge you should be using and acquiring, in the [Transparency in Learning and Teaching (TILT)](tilt.qmd) document in this repository.
The TILT document also contains a checklist for self-reflection that will provide some guidance on how the assignment will be graded.
:::

# Graphical Critiques

The following code was [generated by Claude AI vers. Sonnet 4.5](4-types-graphics-claude.md) (with some modifications).

```{r}
#| label: fig-comparing-graphics-figures
#| fig-cap: "Four different representations of COVID vaccination rates."
#| fig-subcap: ["The Pie Chart shows that over 12 weeks, the leading age range for vaccination was 65+, followed by 50-64, then 30-49, and finally 18-29. This shows that older age groups have more total vaccinations than younger groups.", "The line graph shows that over a 12 week period each age group (18-29, 30-49, 50-64, 60+) all have increases in vaccination rate.The increase seems larger as age range increases with 18-29 being the most shallow and 65+ being the steepest.", "The stacked area graph shows the increase in cumulative vaccination rate over a 12 week period seperate by age group (18-29, 30-49, 50-64, 60+). The graphs shows that each age group has grown in vaccination rate over the period with a total cumalative rate over 260%.", "The bar chart shows that over a 12 week period, vaccination status for each age group increased each week."]
#| fig-alt: ["A pie chart for total vaccinations per age group showing 33.8% for age 65+, 27.7 for 50-64, 21.9 for 30-49, and 16.5 for 18-29.", "A line graph showing differences in vaccination rate for age groups over a 12 week period in which 18-29 increases from 12% to 51%, 30-49 increases from 18% to 64%, 50-64 increases from 25% to 77%, and 65+ increases from 36% to 87%.", "A stacked area chart showing cumulative vaccination rates for age groups (18-29, 30-49, 50-64, 60+) over a 12 week period.", "A bar chart showing differences in vaccination rate for age groups over a 12 week period in which 18-29 increases from 12% to 51%, 30-49 increases from 18% to 64%, 50-64 increases from 25% to 77%, and 65+ increases from 36% to 87%."]
#| layout-ncol: 2
#| code-fold: true


# Load required libraries
suppressPackageStartupMessages({
  library(ggplot2)
  theme_set(theme_bw())
  library(dplyr)
  library(tidyr)
})

# Create sample dataset
set.seed(123)
weeks <- 1:12
age_groups <- c("18-29", "30-49", "50-64", "65+")

# Generate vaccination rate data with realistic patterns
vaccination_data <- data.frame(
  week = rep(weeks, each = 4),
  age_group = rep(age_groups, times = 12),
  vaccination_rate = c(
    # Week-by-week data for each age group
    12, 18, 25, 35,  # Week 1
    15, 22, 30, 41,  # Week 2
    18, 26, 35, 47,  # Week 3
    22, 31, 41, 54,  # Week 4
    26, 36, 47, 60,  # Week 5
    30, 41, 53, 66,  # Week 6
    34, 46, 58, 71,  # Week 7
    38, 50, 63, 75,  # Week 8
    42, 54, 67, 79,  # Week 9
    45, 58, 71, 82,  # Week 10
    48, 61, 74, 85,  # Week 11
    51, 64, 77, 87   # Week 12
  )
)

# Set factor levels for consistent ordering
vaccination_data$age_group <- factor(vaccination_data$age_group, 
                                     levels = c("18-29", "30-49", "50-64", "65+"))

# Define color palette
age_colors <- c("18-29" = "#FF6B6B", 
                "30-49" = "#4ECDC4", 
                "50-64" = "#FFD93D", 
                "65+" = "#95E1D3")

# CHART A: 3D Pie Chart (showing total vaccinations by age group)

# Calculate total vaccinations per age group (sum across weeks)
total_by_age <- vaccination_data %>%
  group_by(age_group) %>%
  summarise(total = sum(vaccination_rate))

# Create pie chart with 3D-like appearance using custom positioning
chart_a <- ggplot(total_by_age, aes(x = "", y = total, fill = age_group)) +
  geom_bar(stat = "identity", width = 1, color = "white") +
  coord_polar("y", start = 0) +
  scale_fill_manual(values = age_colors) +
  labs(title = "Chart A: 3D Pie Chart - Total Vaccinations by Age Group",
       subtitle = "(Summed across all 12 weeks)",
       fill = "Age Group",
       alt = "Test for the alt") +
  theme_void() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 10, face = "italic"),
    legend.position = "right",
    legend.title = element_text(size = 11, face = "bold"),
    legend.text = element_text(size = 10)
  ) +
  # Add percentage labels
  geom_text(aes(label = paste0(round(total/sum(total)*100, 1), "%")),
            position = position_stack(vjust = 0.5),
            size = 4, fontface = "bold")

# CHART B: Line Graph - Vaccination Rates Over Time

chart_b <- ggplot(vaccination_data, aes(x = week, y = vaccination_rate, 
                                        color = age_group, group = age_group)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  scale_color_manual(values = age_colors) +
  scale_x_continuous(breaks = 1:12) +
  labs(title = "Chart B: Line Graph - Vaccination Rates Over Time",
       x = "Week",
       y = "Vaccination Rate (%)",
       color = "Age Group")

# CHART C: Stacked Area Chart - Vaccination Rates Over Time

chart_c <- ggplot(vaccination_data, aes(x = week, y = vaccination_rate, 
                                        fill = age_group)) +
  geom_area(alpha = 0.8, position = "stack") +
  scale_fill_manual(values = age_colors) +
  scale_x_continuous(breaks = 1:12) +
  labs(title = "Chart C: Stacked Area Chart - Vaccination Rates Over Time",
       x = "Week",
       y = "Cumulative Vaccination Rate (%)",
       fill = "Age Group")


# CHART D: Grouped Bar Chart - Vaccination Rates by Week and Age Group

chart_d <- ggplot(vaccination_data, aes(x = factor(week), y = vaccination_rate, 
                                        fill = age_group)) +
  geom_bar(stat = "identity", position = "dodge", width = 0.8) +
  scale_fill_manual(values = age_colors) +
  labs(title = "Chart D: Grouped Bar Chart - Vaccination Rates by Week and Age Group",
       x = "Week",
       y = "Vaccination Rate (%)",
       fill = "Age Group")

chart_a
chart_b
chart_c
chart_d
```

1.  Edit the chunk options above to produce helpful and descriptive captions and alt-text for each chart. Every chart you include in a document should be referenced in the text (preferably using a cross-reference) and should have a helpful caption and include at least short alt-text descriptions using the `fig-alt` quarto option.\
    Resources:

-   [Mastering Alt Text for Images and Charts](https://a11ykat.medium.com/see-the-unseen-mastering-alt-text-for-images-and-charts-6d3dbe36553b)
-   [Image Accessibility Generator](https://asuo-ai-labs.streamlit.app/Image_Accessibility) - This is a great tool but I often have to customize the output. I usually use the long text for image descriptions, which is not *best* practice but tends to work better than many of the other options available within the quarto ecosystem.

2.  Using the grammar of graphics, identify the geom, mappings, stat, positions, coordinate systems, and transformations for each graphic. Fill in the table below accordingly, leaving any cells which do not apply to the chart blank. If a mapping is a function of a statistic, fill in the statistic function from ggplot2 and indicate the mapping using the appropriate [`after_stat` variable](https://ggplot2.tidyverse.org/reference/aes_eval.html.).

| Chart | Geom | X mapping | Y mapping | Color mapping | Fill mapping | Stat | Position | Coord System | Transform |
|-----------|-----------|------------------|-----------------|----------------|---------------|-----------|--------------|-------------|------------------|
| A | Bar | "" | total | X | age group | identity | right | polar | X |
| B | line | time | vaccination rate | age group | X | identity | X | cartesian | X |
| C | area | time | vaccination rate | age group | age group | identity | stack | cartesian | X |
| D | bar | time | vaccination rate | age group | age group | identity | dodge | cartesian | X |

: Grammar of graphics mappings for the four charts

3.  Examine each chart and determine whether it effectively communicates changes in vaccine rates over time across different age groups. If the chart is inappropriate for that comparison, explain why; if it is appropriate but not optimal, explain what you would change.


    A. No, there is no way to tell the difference within or between groups over 12 weeks with this graph. We can only see the final total for each group. My only idea on how to make this style of graph work would be to make 12 pie charts 
    
    B.  This graph does a better job showing the intended information. Out of all the graphs this is the optimal graph. We can see the percent for each group at each week easily and can see the differences between groups.
    
    C.  This graph is not good to show the changes in vaccine rates for the groups. We can see the changes, it does work to display the data, but not well. Its not optimal because the cumulative rate goes above 100% which is confusing becasue it adds up each group. Its hard to tell any information from this graph.
    
    D.  This graph does show the information we need for the differences between groups over the 12 weeks. However, the bar charts make the graph more difficult to understand and is needless. The line chart does what this graph does much better.


4.  Examine each chart and critique the use of Gestalt principles of grouping. What principles apply? Are they used effectively?

    
    A.  The colors for each group are different for similarity. There are slices between each piece of the pie so its not very good with closure. Its decent with gestalt
    
    B.  There is good continuity with the lines in the graph. There is good similarity with the lines and points and proximity becasue they are all on the line. I would say this is effective
    
    C.  There is continuity with the areas. There is proximity because the lines are all connected. Closure for each band being a solid object. This is decent with gestalt. 
    
    D.  Each weekly grouping is closer to eachother than other weeks for good proximity, there is similarity between the group colors. I would say this one is not very good though because it feels too noisy.


5.  Consider the "messages" you might want viewers to take away from this data and the visuals you generate.

    a.  Brainstorm 2-3 reasonable messages    
    
    > I think the messages should be that vaccination rate steadily rose over 12 weeks and that the older a group is the higher the vaccine rate.
    
    b.  What do you think the most important message is from this data set?    
    
    > The most important message is that all groups raised in vaccination rate over time
    
    c.  Which comparisons do you want to make easier for participants to perform?    
    
    > Week by week comparisons for each group and comparisons for what the actual percentage each point is
    
    d.  What obstacles do you expect to encounter designing a chart to facilitate these comparisons?    
    
    > Finding good colors, I want to differentiate the older groups from younger


6.  Considering your answers to the questions above, create your own version of the chart that minimizes cognitive load and maximizes interpretability. Explain the changes that you made and why they improve the chart compared to the ones generated by Claude.

I changed the colors to be more distinguishable as well as made the y axis increase by 10s to make it easier to see the value of each point. Both these changes help to reduce the cognitive load of finding a value.

```{r}

age_colors <- c("18-29" = "#3594cc", 
                "30-49" = "#8cc5e3", 
                "50-64" = "#ea801c", 
                "65+" = "#f0b077")


chart_b <- ggplot(vaccination_data, aes(x = week, y = vaccination_rate, color = age_group, group = age_group)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  scale_color_manual(values = age_colors) +
  scale_x_continuous(breaks = seq(1, 12, by=1)) +
  scale_y_continuous(breaks = seq(0, 100, by = 10)) +
  labs(title = "Vaccination Rates Over Time",
       x = "Week",
       y = "Vaccination Rate (%)",
       color = "Age Group")

chart_b

```



# Choose Your Own Adventure

Claude used fake data, but we have access to the [real deal](https://data.cdc.gov/Vaccinations/Weekly-Cumulative-Estimated-Number-of-COVID-19-Vac/ewpg-rz7g/about_data), and I've downloaded a [csv](Weekly_Cumulative_Estimated_Number_of_COVID-19_Vaccinations_Administered_in_Pharmacies_and_Physician_Medical_Offices,_Adults_18_years_and_older,_by_Season_and_Age_Group,_United_States_20251021.csv) to simplify the process.

Of course, the CDC only tracks number of vaccines, not the overall vaccination rate.

1.  Augment this data with census estimates of the population size for each age group. 
In either case, you may need to do a bit of inference/subtraction to get the counts for the age brackets used by the CDC.
Provide reproducible code that generates a population table for 2023 and 2024 with the appropriate counts for each age bracket provided by the CDC.    
    
    Options:
    
    -   Use the Census QuickFacts page https://www.census.gov/quickfacts/
    -   Use the [nc-est2024-agesex.xlsx](nc-est2024-agesex.xlsx) population by age and sex file I downloaded from the [US Census](https://www.census.gov/data/tables/time-series/demo/popest/2020s-national-detail.html).
    
    
```{r}
library(readr)
library(dplyr)

vax_data <- read_csv("Weekly_Cumulative_Estimated_20251021.csv") |>
  filter(Setting == "Combined", Age_Group %in% c("18-49","50-64","65+","18+")) |>
  select(Week_ID, Age_Group, Cumulative_Doses)

pop_data <- read_csv("nc-est2024-agesex.csv")[, c(1,17)]
names(pop_data) <- c("label","value")
pop_data$value <- as.numeric(gsub(",", "", pop_data$value))

p <- setNames(pop_data$value, pop_data$label)


pop_table <- tibble(
  Age_Group = c("18-49","50-64","65+","18+"),
  Population = c(p[".18 to 24 years"] + p[".25 to 44 years"] + p[".45 to 49 years"], p[".50 to 54 years"] + p[".55 to 59 years"] + p[".60 to 64 years"], p[".65 years and over"], p[".18 years and over"])
)

final <- vax_data |>
  inner_join(pop_table, by = "Age_Group") |>
  mutate(share_of_population_pct = 100 * Cumulative_Doses / Population)

```

    

2.  Use the real data to generate the plot you created in 6. How does the real data change the appearance/message of the plot? Does anything get easier or more difficult?

the real data only has 3 age groups so that changes alot. The message is much clearer that older people got vaccinated faster. It was difficult to get the data in the right form to plot.


```{r}
library(dplyr)
library(ggplot2)

fmt <- "%Y %b %d %I:%M:%S %p"
week_levels <- unique(final$Week_ID[order(as.POSIXct(final$Week_ID, format = fmt, tz = "UTC"))])
first12 <- head(week_levels, 12)

week_map <- tibble(Week_ID = first12, week_idx = 1:12)
plot_df <- final %>% 
  filter(Week_ID %in% first12, Age_Group != "18+") %>%
  inner_join(week_map, by = "Week_ID")

age_cols <- c("18-49" = "#8cc5e3",
              "50-64" = "#ea801c",
              "65+"   = "#f0b077")

ggplot(plot_df, aes(x = week_idx, y = share_of_population_pct, color = Age_Group, group = Age_Group)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  scale_color_manual(values = age_cols, name = "Age Group") +
  scale_x_continuous(breaks = 1:12, labels = paste0(1:12)) +
  labs(title = "Vaccination Rates (First 12 Weeks, 18+ Removed)",
       x = "", y = "Percent of Population Vaccinated") +
  scale_y_continuous(labels = function(x) paste0(x, "%"))


```



3.  What annotations would you want to add to your chart? You may be able to find relevant information about e.g. approval dates of vaccines at [immunize.org's Vaccine Timeline](https://www.immunize.org/vaccines/vaccine-timeline/). Add at least one annotation to your chart, but balance utility with clutter.

```{r}
library(dplyr)
library(ggplot2)

fmt <- "%Y %b %d %I:%M:%S %p"
week_levels <- unique(final$Week_ID[order(as.POSIXct(final$Week_ID, format = fmt, tz = "UTC"))])
first12 <- head(week_levels, 12)

week_map <- tibble(Week_ID = first12, week_idx = 1:12)
plot_df <- final %>% 
  filter(Week_ID %in% first12, Age_Group != "18+") %>%
  inner_join(week_map, by = "Week_ID")

age_cols <- c("18-49" = "#8cc5e3",
              "50-64" = "#ea801c",
              "65+"   = "#f0b077")

ggplot(plot_df, aes(x = week_idx, y = share_of_population_pct, color = Age_Group, group = Age_Group)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  scale_color_manual(values = age_cols, name = "Age Group") +
  scale_x_continuous(breaks = 1:12, labels = paste0(1:12)) +
  labs(title = "Vaccination Rates (First 12 Weeks, 18+ Removed)",
       x = "", y = "Percent of Population Vaccinated") +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  annotate("text", x = 2, y = max(plot_df$share_of_population_pct), label = "Older age \u2192 higher COVID-19 fatal risk", hjust = 0, size = 3.5)
```


4.  How would you modify this chart if your goal was to compare vaccination rates for each age group in 2023 and 2024?
    Explain your answer, highlighting how your modifications would reduce the cognitive load of making the comparison of interest.

 I would make it so instead of just plotting 12 weeks it plots the entire years. This would probably make the graph alot more dense so I would need to scale it down and change the axis to plot maybe every month or so.

5.  How would you use your statistical training to test and visually highlight any statistically significant differences?
    Explain.
    You do not have to actually implement any statistical testing to answer this question.
    
    I think it would be interesting to see if there was any weeks were there was a significant difference in the rate of change of vaccination rate. We could test this to see which if any weeks apply, then potentially make an annotation or change the color of that week to show that.
